---
output: github_document
---

```{r echo=FALSE, results = 'asis'}
pkg <- 'seriation'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```


This package provides the infrastructure for ordering objects 
with an implementation of several
[seriation](https://en.wikipedia.org/wiki/Seriation_(archaeology))/sequencing/[ordination](https://en.wikipedia.org/wiki/Ordination_(statistics)) techniques to reorder matrices, dissimilarity
matrices, and dendrograms (see below for a full list). Also provides (optimally) reordered heatmaps, 
color images and clustering visualizations like dissimilarity plots, and
visual assessment of cluster tendency plots (VAT and iVAT).

## Available Seriation Method

The following methods are available for dissimilarity data:

 *  **ARSA** - Simulated annealing (linear seriation)   
 *  **Branch-and-bound** to minimize the unweighted/weighted column gradient 
 *  **DendSer** - Dendrogram seriation heuristic to optimize various criteria
 *  **GA** - Genetic algorithm with warm start to optimize various criteria
 *  **GW** - Hierarchical clustering reordered by Gruvaeus and Wainer heuristic 
 *  **HC** - Hierarchical clustering (single link, avg. link, complete link) 
 *  **Identity** permutation 
 *  **MDS** - Multidimensional scaling (metric, non-metric, angle) 
 *  **OLO** - Hierarchical clustering with optimal leaf ordering 
 *  **OPTICS** - Ordering points to identify the clustering structure.
 *  **QAP** - Quadratic assignment problem heuristic (2-SUM, linear seriation, inertia, banded anti-Robinson form)
 *  **R2E** - Rank-two ellipse seriation 
 *  **Random** permutation
 *  **Spectral** seriation (unnormalized, normalized) 
 *  **SPIN** - Sorting points into neighborhoods (neighborhood algorithm, side-to-site algorithm)
 *  **TSP** - Traveling sales person solver to minimize the Hamiltonian path length 
 *  **TSNE** - Order of the 1D t-distributed stochastic neighbor embedding (t-SNE)
 *  **UMAP** - Order of the 1D embedding produced by uniform manifold approximation and projection
 *  **VAT** - Order of the visual assessment of clustering tendency ordering 
  
A detailed comparison of the methods is available in the paper 
[An experimental comparison of seriation methods for one-mode two-way data.](http://dx.doi.org/10.1016/j.ejor.2016.08.066) (read the [preprint](https://michael.hahsler.net/research/paper/EJOR_seriation_2016.pdf)).
  
  
The following methods are available for matrices:

 *  **BEA** - Bond Energy Algorithm to maximize the measure of effectiveness (ME) 
 *  **Identity** permutation 
 *  **PCA** - First principal component or angle on the projection on the first two principal components 
 *  **Random** permutation 
 *  **TSP** - Traveling sales person solver to maximize ME 


```{r echo=FALSE, results = 'asis'}
pkg_install(pkg)
```

## Usage

Load library, read data and calculate distances. Then use default seriation.
```{r}
library(seriation)
data("iris")
x <- as.matrix(iris[-5])
x <- x[sample(1:nrow(x)),]

d <- dist(x)
order <- seriate(d)
order
```
```{r seriation}
pimage(d, main = "random")
pimage(d, order, main = "reordered")
```



Compare quality.
```{r}
rbind(
 random = criterion(d),
 reordered = criterion(d, order)
)
```


## References

* Michael Hahsler, Kurt Hornik and Christian Buchta, [Getting Things in Order: An Introduction to the R Package seriation,](http://dx.doi.org/10.18637/jss.v025.i03) _Journal of Statistical Software,_ 25(3), 2008.
* Michael Hahsler. [An experimental comparison of seriation methods for one-mode two-way data.](http://dx.doi.org/10.1016/j.ejor.2016.08.066) _European Journal of Operational Research,_ 257:133-143, 2017. (read the [preprint](https://michael.hahsler.net/research/paper/EJOR_seriation_2016.pdf))
* [Reference manual for package seriation.](https://www.rdocumentation.org/packages/seriation/)
* [Seriation package vignette](https://cran.r-project.org/package=seriation/vignettes/seriation.pdf) with complete examples.

