---
output: github_document
---

```{r echo=FALSE, results = 'asis'}
pkg <- 'seriation'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```


This package provides the infrastructure for ordering objects 
with an implementation of several
[seriation](https://en.wikipedia.org/wiki/Seriation_(archaeology))/sequencing/[ordination](https://en.wikipedia.org/wiki/Ordination_(statistics)) techniques to reorder matrices, dissimilarity
matrices, and dendrograms (see below for a full list). Also provides (optimally) reordered heatmaps, 
color images and clustering visualizations like dissimilarity plots, and
visual assessment of cluster tendency plots (VAT and iVAT).

## Available Seriation Method

The following methods are available to reorder dissimilarity data:

 **Optimization-based**
 
 *  **ARSA** - optimize the linear seriation critreion using simulated annealing   
 *  **Branch-and-bound** to minimize the unweighted/weighted column gradient 
 *  **GA** - Genetic algorithm with warm start to optimize any seriation criteria
 *  **GSA** - General simmulated annealing to optimize any seriation criteria
 *  **QAP** - Quadratic assignment problem heuristic (2-SUM, linear seriation, 
          inertia, banded anti-Robinson form)
 *  **Spectral** seriation (unnormalized, normalized) 
 *  **TSP** - Traveling sales person solver to minimize the Hamiltonian path length 
 
**Multidimensional Scaling** 

 *  **MDS** - classical metric multidimensional scaling
 *  **Angle in 2D principal coordinates space** 
 *  **isoMDS** -  Krusakl's isoMDS
 *  **isomap** -  Isometric feature mapping ordination
 *  **monoMDS** - Global and local non-metric multidimensional scaling (NMDS).
 *  **metaMDS** - Nonmetric multidimensional scaling with stable solution from random starts
 *  **Sammon** - Order along the 1D Sammon's non-linear mapping
 *  **smacof** - Stress minimization using majorization (metric/nonmetric)
 *  **TSNE** - Order along the 1D t-distributed stochastic neighbor embedding (t-SNE)
 *  **UMAP** - Order along the 1D embedding produced by uniform manifold approximation and projection
 *  **VAE** - Order along the 1D embedding produced by a variational autoencoder.
 
 **Dendrogram-based**
 
 *  **DendSer** - Dendrogram seriation heuristic to optimize various criteria
 *  **GW** - Hierarchical clustering reordered by the Gruvaeus and Wainer heuristic 
 *  **HC** - Hierarchical clustering (single link, avg. link, complete link) 
 *  **OLO** - Hierarchical clustering with optimal leaf ordering 
 
 **Other**

 *  **Identity** permutation 
 *  **OPTICS** - Ordering points to identify the clustering structure.
 *  **R2E** - Rank-two ellipse seriation 
 *  **Random** permutation
 *  **Reverse** order
 *  **SPIN** - Sorting points into neighborhoods (neighborhood algorithm, side-to-site algorithm)
 *  **VAT** - Order of the visual assessment of clustering tendency
  
A detailed comparison of the methods is available in the paper 
[An experimental comparison of seriation methods for one-mode two-way data.](http://dx.doi.org/10.1016/j.ejor.2016.08.066) (read the [preprint](https://michael.hahsler.net/research/paper/EJOR_seriation_2016.pdf)).
  
  
The following methods are available to reorder rows and columns of data matrices:

 *  **BEA** - Bond Energy Algorithm to maximize the measure of effectiveness (ME) 
 *  **BEA_TSP** - TSP to optimize the Measure of Effectiveness
 *  **CA** - calculates a correspondence analysis of a matrix of frequencies (count table)
      and reorders according to the scores on a correspondence analysis dimension
 *  **Heatmap** - reorders rows and columns independently by calculation row/column distances 
      and applying a seriation method (default is optimal leaf ordering) 
 *  **Identity** permutation 
 *  **LLE** reorder along a 1D locally linear embedding. 
 *  **Means** - reorders using row and column means.
 *  **PCA** - orders along the first principal component or angle on the projection on the first two principal components
 *  **Random** permutation 
 *  **Reverse** order
 *  **TSNE** - Order along the 1D t-distributed stochastic neighbor embedding (t-SNE)
 *  **UMAP** - Order along the 1D embedding produced by uniform manifold approximation and projection
 *  **VAE** - Order along the 1D embedding produced by a variational autoencoder.

```{r echo=FALSE, results = 'asis'}
pkg_install(pkg)
```

## Usage

Load library, read data and calculate distances. Then use default seriation. The used dataset 
contains the joint probability of disagreement between Supreme Court Judges during the period
from 1995 to 2002.
```{r}
library(seriation)
data("SupremeCourt")

SupremeCourt

d <- as.dist(SupremeCourt)
order <- seriate(d)
order
```

Get the resulting permutation vector.
```{r}
get_order(order)
```

```{r seriation,  fig.show="hold", out.width="50%"}
pimage(d, diag = TRUE, upper = TRUE, main = "judges in original alphabetical order")
pimage(d, order, diag = TRUE, upper = TRUE, main = "judges reordered by seriation")
```



Compare several available quality metrics.
```{r}
rbind(
 alphabetical = criterion(d),
 seriated = criterion(d, order)
)
```

## References

* Michael Hahsler, Kurt Hornik and Christian Buchta, [Getting Things in Order: An Introduction to the R Package seriation,](http://dx.doi.org/10.18637/jss.v025.i03) _Journal of Statistical Software,_ 25(3), 2008.
* Michael Hahsler. [An experimental comparison of seriation methods for one-mode two-way data.](http://dx.doi.org/10.1016/j.ejor.2016.08.066) _European Journal of Operational Research,_ 257:133-143, 2017. (read the [preprint](https://michael.hahsler.net/research/paper/EJOR_seriation_2016.pdf))
* [Reference manual for package seriation.](https://www.rdocumentation.org/packages/seriation/)

