---
title: "Getting started with seriation"
author: "Michael Hahsler"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with seriation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
library(seriation)
set.seed(1234)
```

Seriation arranges objects in a linear order so that related objects are close
to one another. Reordering does not change the data; it changes only how rows,
columns, or objects are presented. A useful order can expose clusters,
gradients, and other structure that is difficult to see in the original
ordering.

Package `seriation` provides a common interface to many seriation methods and
tools for applying, assessing, and visualizing the resulting orders. This
vignette introduces how to use seriation for distance and data matrices,
and how to assess the quality of an resulting order. Finally, some guidance for
choosing an appropriate method are given.

## Installation

Install the released version from CRAN:

```{r install, eval=FALSE}
install.packages("seriation")
```

Load the package in each R session where you want to use it:

```{r load-package}
library(seriation)
```

## Seriate a dissimilarity matrix

We first use the `SupremeCourt` data. It contains pairwise disagreement
probabilities for nine U.S. Supreme Court justices. Because these values are
dissimilarities, we convert the symmetric matrix to an R `dist` object.

```{r prepare-distance}
data("SupremeCourt")
d <- as.dist(SupremeCourt)
d
```

The original rows and columns are alphabetical. A permutation image plot
shows the dissimilarities using color; darker cells represent smaller
dissimilarities.

```{r original-distance, fig.width=5, fig.height=5}
pimage(d, main = "Original alphabetical order")
```

`seriate()` computes an order. Methods are selected by name. Here we use
spectral seriation, which is also the default for a `dist` object.

```{r find-order}
o <- seriate(d, method = "Spectral")
o
```

The result is a `ser_permutation` object. `get_order()` extracts the ordinary
integer permutation vector, with names showing the labels in their new order.

```{r inspect-order}
get_order(o)
```

The vector can be used for standard R subsetting. More generally,
`permute()` applies a seriation order while preserving the input object's
class.

```{r apply-order}
d_ordered <- permute(d, o)
as.matrix(d_ordered)[1:4, 1:4]
```

Most plotting functions in the package accept the order directly, so it is
usually unnecessary to create a reordered copy just for visualization.

```{r reordered-distance, fig.width=5, fig.height=5}
pimage(d, order = o, main = "Spectral seriation")
```

The reordered plot places justices with similar voting patterns next to one
another and reveals two darker blocks along the diagonal.

## Assess the quality of a seriation order

`criterion()` calculates objective measures for an order. The spectral method
targets the 2-Sum criterion. Both 2-Sum and Hamiltonian path length are loss
functions, so smaller values are better.

```{r assess-order}
rbind(
  original = criterion(d, method = c("2SUM", "Path_length")),
  seriated = criterion(d, o, method = c("2SUM", "Path_length"))
)
```

Different methods optimize different criteria, and a method that performs
well for one goal need not be best for another. Use `criterion()` to compare
orders using measures that match the purpose of the analysis.

## Seriate a data matrix

For a rectangular matrix, rows and columns can be seriated separately. The
`Wood` data contains expression measurements for 136 genes at six locations.

```{r matrix-order}
data("Wood")
dim(Wood)

o_matrix <- seriate(Wood, method = "Heatmap")
o_matrix
```

This order contains one permutation for each matrix dimension. Use `dim = 1`
for rows and `dim = 2` for columns.

```{r matrix-permutations}
head(get_order(o_matrix, dim = 1))
get_order(o_matrix, dim = 2)
```

The same object can reorder both dimensions with `permute()` or can be passed
directly to `pimage()`.

```{r matrix-images, fig.show="hold", out.width="49%", fig.width=5, fig.height=5}
pimage(Wood, main = "Original order")
pimage(Wood, order = o_matrix, main = "Seriated rows and columns")
```

To order only one dimension, use the `margin` argument. For example, this
orders rows using the first principal component and leaves columns unchanged:

```{r one-margin}
o_rows <- seriate(Wood, method = "PCA", margin = 1)
head(get_order(o_rows, dim = 1))
```

## Choose a method

Available methods depend on the input type. Use
`list_seriation_methods()` to discover valid method names. 

```{r methods}
list_seriation_methods("dist")
list_seriation_methods("matrix")
```

A comprehensive comparison of different seriation methods can be found in
the paper 
[An experimental comparison of seriation methods for one-mode two-way data](http://dx.doi.org/10.1016/j.ejor.2016.08.066)
([open-access preprint](http://michael.hahsler.net/research/paper/EJOR_seriation_2016.pdf)).
Useful starting points are:

- `"Spectral"` and `"QAP_2SUM"` for general dissimilarity data,
- `"OLO"` for a hierarchical clustering with optimal leaf ordering,
- `"TSP"` for a short path through the objects, and
- `"Heatmap"` for ordering the rows and columns of a data matrix using
  separate dissimilarities.
  

Detailed method descriptions, what measure is optimized, 
and a list with available control parameters can be obtained 
with `get_seriation_method()`.

```{r methods2}
get_seriation_method("dist", "Spectral")
```

Some methods are randomized. For those methods, set a seed for reproducibility
and use the `rep` argument to keep the best result from several restarts.
Method-specific arguments can be supplied to `seriate()` in `control` or 
directly through `...`.

## Where to go next

The package includes more focused guides:

- [Seriation methods](seriation_methods.html) lists the available algorithms
  and their control parameters.
- [Seriation criteria](seriation_criteria.html) describes measures for
  evaluating an order.
- [Heatmaps](heatmaps.html) covers `pimage()`, `hmap()`, and `gghmap()`.
- [Correlation matrices](correlation_matrix.html) shows methods designed for
  correlations.
- [Seriation and clustering](clustering.html) demonstrates reordered
  dissimilarity plots and cluster visualization.

The help pages `?seriate`, `?permute`, `?criterion`, and `?pimage` provide the
complete function reference.
