---
title: "Heatmaps with Package Seriation"
author: "Michael Hahsler"
format: markdown_github
editor: visual
---

The package seriation can be used for reordering heatmaps and displaying them
using the functions

* `hmap()`
* `gghmap()`

There functions can be used as a direct replacement of `heatmap()` in the 
R package `stats`.

## Standard Heatmaps

Here are some typical reordering schemes.

```{r}
library("seriation")
data("Wood")
```

```{r}
hmap(Wood, method = "HC", main = "Wood (hierarichal clustering; no reordering)")
hmap(Wood, method = "GW", main = "Wood (reordered using the Gruvaeus and Wainer heuristic)")
hmap(Wood, method = "OLO", main = "Wood (opt. leaf ordering)")
```

Different linkage types can be added in the method name.

Package `DendSer` offers more dendrogram seriation methods. These methods can be 
registered using `register_DendSer()``

```{r}
register_DendSer()

hmap(Wood, method = "DendSer_BAR", main = "Wood (banded anti-Robinson)")
```

## Non-dendrogram-based seriation methods

These methods reorder rows and columns. No dendrogram is produced.

```{r}
hmap(Wood, method = "MDS", main = "Wood (MDS)")
hmap(Wood, method = "Spectral", main = "Wood (MDS)")
```

## ggplot2 version

All options are also available for `ggplot2` using `gghmap()`. 
Currently there is no support to display dendrograms.

```{r}
library(ggplot2)
gghmap(Wood, method = "OLO")
```

Package `heatmaply` uses `seriation` to reorder interactive heatmaps.

