---
title: "Correlation Matix Visualization with Package Seriation"
author: "Michael Hahsler"
format:
  html:
    toc: true
editor: visual
---

# Introduction

A correlation matrix is a square, symmetric matrix showing the pairwise 
correlation coefficients between two sets of variables. Reordering 
the variables and plotting the matrix can help to find hidden patterns among 
the variables. The package `seriation` implements a large number of reordering methods (see: the [list with all implemented seriation methods](https://mhahsler.github.io/seriation/seriation_methods.html)). `seriation` also provides a set of functions to display reordered matrices:

-   `pimage()`
-   `ggpimage()`

How to cite the `seriation` package:

> Hahsler M, Hornik K, Buchta C (2008). "Getting things in order: An introduction to the R package seriation." *Journal of Statistical Software*, *25*(3), 1-34. ISSN 1548-7660, doi:10.18637/jss.v025.i03 <https://doi.org/10.18637/jss.v025.i03>.

## Prepare the data

As an example, we use the `mtcars` dataset 
which contains data about fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

```{r}
if (!require("seriation")) install.packages("seriation")

library("seriation")
data("mtcars")

DT::datatable(mtcars)
```

We calcualte a correlation matrix.

```{r}
m <- cor(mtcars)
round(m, 2)
```

We first visualize the matrix without reordering and then use the order method `"AOE"`.
AOE stands for angle of eigenvectors and was proposed for correlation matrices by
Friendly (2002).


```{r}
#| fig-asp: 1
#| fig-show: hold
#| out-width: 50%
#| layout-ncol: 2 
pimage(m)
pimage(m, order = "AOE")
```

The reordering clearly shows that there is tow groups of highly correlated variables and these
two groups have a strong negative correlation with each other.


## Visualization options

Here are some options.

```{r}
#| fig-asp: 1
#| fig-show: hold
#| out-width: 50%
#| layout-ncol: 2 
pimage(m, order = "AOE", diag = FALSE, upper_tri = FALSE)
pimage(m, order = "AOE", col = colorRampPalette(c("red", "white", "darkgreen"))(100))
```


The plots are also available in `ggplot2` versions.

```{r}
#| fig-asp: 1
#| fig-show: hold
#| out-width: 50%
#| layout-ncol: 2 
library("ggplot2")

ggpimage(m, order = "AOE", diag = FALSE, upper_tri = FALSE)
ggpimage(m, order = "AOE") + scale_fill_gradient2(low = "red", high = "darkgreen")
```



## Using other seriation methods

We can apply any seriation method for distances to create an order.
First, we convert the correlation matrix into a distance matrix using
$d_{ij} = \sqrt{1 - m_{ij}}$. Then we can use the distances for seriation
and use the resulting order to rearrange the rows and columns of the 
correlation matrix.

```{r}
#| fig-asp: 1
#| fig-show: hold
#| out-width: 50%
#| layout-ncol: 2 
d <- as.dist(sqrt(1 - m))

o <- seriate(d, "MDS")
pimage(m , order = c(o, o), main = "MDS")

o <- seriate(d, "ARSA")
pimage(m , order = c(o, o), main = "ARSA")

o <- seriate(d, "OLO")
pimage(m , order = c(o, o), main = "OLO")

o <- seriate(d, "R2E")
pimage(m , order = c(o, o), main = "R2E")
```


## Other packages

Several other packages can be used to visualize and explore correlation structure.
Some of these packages support reordering with the seriation package.

### Package corrgram


The order argument in [corrgram](https://kwstat.github.io/corrgram/) accepts methods from package seriation.

```{r}
#| fig-asp: 1
#| fig-show: hold
#| out-width: 50%
#| layout-ncol: 2 
if (!require("corrgram")) install.packages("corrgram")
library("corrgram")

corrgram(m, order = "OLO")
corrgram(m, order = "OLO", lower.panel=panel.shade, upper.panel=panel.pie)
```

### Package corrr

The function `rearrange()` in package [corrr](https://corrr.tidymodels.org/) accepts some methods from seriation. Here is a 
complete example that uses method `"R2E"`.

```{r}
#| fig-asp: 1
#| out-width: 50%
#| fig-align: "center"
if (!require("corrr")) install.packages("corrr")
library("corrr")

x <- datasets::mtcars |>
       correlate() |>   
       focus(-cyl, -vs, mirror = TRUE) |>  # remove 'cyl' and 'vs'
       rearrange(method = "R2E") |>  
       shave()

rplot(x)
```


### Package corrplot

Package [corrplot](https://github.com/taiyun/corrplot) offers many visualization 
methods. Orders from package seriation can be used by permuting 
the correlation matrix before it is passed to `corrplot()`.

```{r}
#| fig-asp: 1
#| out-width: 50%
#| fig-align: "center"
if (!require("corrplot")) install.packages("corrplot")
library("corrplot")

d <- as.dist(sqrt(1 - m))
o <- seriate(d, "R2E")
m_R2E <- permute(m, c(o,o))

corrplot(m_R2E , order = "original")
```

## References

* Michael Hahsler, Kurt Hornik and Christian Buchta, [Getting Things in Order: An Introduction to the R Package seriation,](http://dx.doi.org/10.18637/jss.v025.i03) _Journal of Statistical Software,_ 25(3), 2008.
DOI: 10.18637/jss.v025.i03
* Friendly, M. (2002): Corrgrams: Exploratory Displays for Correlation
Matrices. \emph{The American Statistician}, \bold{56}(4), 316--324.
DOI: 10.1198/000313002533

