---
title: "List of Seriation Methods implemented in the R package seriation"
author: "Michael Hahsler"
format: 
  html:
    toc: true
editor: visual
---

This document contains the seriation methods currently implemented in the R package [`seriation`](https://github.com/mhahsler/seriation). The methods are organized by methods for distance matrices and data matrices.

This list was created for the following version of seriation:

```{r}
library("seriation")
packageVersion('seriation')
```

Some additional methods need to be registered.

```{r, results="hide", warning=FALSE, message=FALSE}
register_DendSer()
register_optics()
register_smacof()
register_GA()
register_tsne()
register_umap()
```

```{r, echo = FALSE}
print_seriation_method_md <- function(x, add_link = TRUE, ...) {
    cat("\n\n###", x$name, "\n")
  
    cat(x$description, "\n\n")
    
#    cat("* kind:", x$kind, "\n")
    
    optimizes <- x$optimizes
    
    .make_anchor <- function(x) {
      # remove leading numbers
      x <- gsub("^\\d+", "", x)
      # lower case
      tolower(x)
    }
    
    opt_info <- attr(optimizes, "description")
    if(is.na(optimizes)) {
      optimizes <- "N/A"
    } else {
      
      if (add_link) {
        optimizes <- paste0("[", optimizes, "]", "(seriation_criteria.html#", .make_anchor(optimizes) ,")")
      }
   
    }
      
    if(!is.null(opt_info)) {
        optimizes <- paste0(optimizes, " (", opt_info, ")")
    }
     
    cat("* optimizes:", optimizes, "\n")
    cat("* randomized:", x$randomized, "\n")
    
    if (!is.na(x$registered_by)) {
      cat("* registered by: ", x$registered_by, "\n")
    }

  cat("* control parameters: ")
  .print_control(x$control, trim_values = 100)

  invisible(x)
}


.print_control <- function(control,
                           help = TRUE,
                           trim_values = 30L) {
  if (length(control) < 1L) {
    writeLines("no parameters")
  } else{
    contr <- lapply(
      control,
      FUN = function(x)
        strtrim(paste(deparse(x), collapse = ""), trim_values)
    )

    contr <- as.data.frame(t(as.data.frame(contr)))
    colnames(contr) <- c("default")

    contr <- cbind(contr, help = "N/A")
    if (!is.null(attr(control, "help")))
      for (i in seq(nrow(contr))) {
        hlp <- attr(control, "help")[[rownames(contr)[i]]]
        if (!is.null(hlp))
        contr[["help"]][i] <- hlp
      }
    print(knitr::kable(contr))
  }

}
```

Seriation methods on the data `x` can be used in

```{r, eval=FALSE}
seriate(x, method = "Spectral", control = NULL, rep = 1L)
```

The list below shows what seriation criterion (if any) is optimized by the method.

If the method uses randomization (see randomized below), then `rep` can be used to run the algorithms several times and return the best permutation.

Available control parameters can be passed on as a list using the `control` argument.

## Methods for dissimilarity data (dist)

```{r, results='asis', echo = FALSE}
l <- list_seriation_methods(kind = "dist", names_only = FALSE)
for(i in seq_along(l))
  print_seriation_method_md(l[[i]])
```

## Methods for data matrices, tables and data.frames

```{r, results='asis', echo = FALSE}
l <- list_seriation_methods(kind = "matrix", names_only = FALSE)
for(i in seq_along(l))
  print_seriation_method_md(l[[i]])
```

# References

Arabie, P. and L.J. Hubert (1990): The bond energy algorithm revisited, *IEEE Transactions on Systems, Man, and Cybernetics,* **20**(1), 268--274. <https://doi.org/10.1109/21.47829>

Bar-Joseph, Z., E. D. Demaine, D. K. Gifford, and T. Jaakkola. (2001): Fast Optimal Leaf Ordering for Hierarchical Clustering. *Bioinformatics,* **17**(1), 22--29. <https://doi.org/10.1093/bioinformatics/17.suppl_1.S22>

Barnard, S. T., A. Pothen, and H. D. Simon (1993): A Spectral Algorithm for Envelope Reduction of Sparse Matrices. *In Proceedings of the 1993 ACM/IEEE Conference on Supercomputing,* 493--502. Supercomputing '93. New York, NY, USA: ACM. \url{https://ieeexplore.ieee.org/document/1263497}

Bezdek, J.C. and Hathaway, R.J. (2002): VAT: a tool for visual assessment of (cluster) tendency. *Proceedings of the 2002 International Joint Conference on Neural Networks (IJCNN '02),* Volume: 3, 2225--2230. <https://doi.org/10.1109/IJCNN.2002.1007487>

Brusco, M., Koehn, H.F., and Stahl, S. (2008): Heuristic Implementation of Dynamic Programming for Matrix Permutation Problems in Combinatorial Data Analysis. *Psychometrika,* **73**(3), 503--522. <https://doi.org/10.1007/s11336-007-9049-5>

Brusco, M., and Stahl, S. (2005): *Branch-and-Bound Applications in Combinatorial Data Analysis.* New York: Springer. <https://doi.org/10.1007/0-387-28810-4>

Chen, C. H. (2002): Generalized Association Plots: Information Visualization via Iteratively Generated Correlation Matrices. *Statistica Sinica,* **12**(1), 7--29.

Ding, C. and Xiaofeng He (2004): Linearized cluster assignment via spectral ordering. *Proceedings of the Twenty-first International Conference on Machine learning (ICML '04)*. <https://doi.org/10.1145/1015330.1015407>

Climer, S. and Xiongnu Zhang (2006): Rearrangement Clustering: Pitfalls, Remedies, and Applications, *Journal of Machine Learning Research,* **7**(Jun), 919--943.

D. Earle, C. B. Hurley (2015): Advances in dendrogram seriation for application to visualization. *Journal of Computational and Graphical Statistics,* **24**(1), 1--25.

Friendly, M. (2002): Corrgrams: Exploratory Displays for Correlation Matrices. *The American Statistician,* **56**(4), 316--324. <https://doi.org/10.1198/000313002533>

Friendly, M. (2023). _vcdExtra: 'vcd' Extensions and Additions_. R package version 0.8-5, <https://CRAN.R-project.org/package=vcdExtra>.

Gruvaeus, G. and Wainer, H. (1972): Two Additions to Hierarchical Cluster Analysis, *British Journal of Mathematical and Statistical Psychology,* **25**, 200--206. <https://doi.org/10.1111/j.2044-8317.1972.tb00491.x>

Hahsler, M. (2017): An experimental comparison of seriation methods for one-mode two-way data. *European Journal of Operational Research,* **257**, 133--143. <https://doi.org/10.1016/j.ejor.2016.08.066>

Hubert, Lawrence, and James Schultz (1976): Quadratic Assignment as a General Data Analysis Strategy. *British Journal of Mathematical and Statistical Psychology,* **29**(2). Blackwell Publishing Ltd. 190--241. <https://doi.org/10.1111/j.2044-8317.1976.tb00714.x>

Hurley, Catherine B. (2004): Clustering Visualizations of Multidimensional Data. *Journal of Computational and Graphical Statistics,* **13**(4), 788--806. <https://doi.org/10.1198/106186004X12425>

Kruskal, J.B. (1964). Nonmetric multidimensional scaling: a numerical method. *Psychometrika,* **29**, 115--129.

Lenstra, J.K (1974): Clustering a Data Array and the Traveling-Salesman Problem, *Operations Research,* **22**(2) 413--414. <https://doi.org/10.1287/opre.22.2.413>

Mair P., De Leeuw J. (2015). Unidimensional scaling. In *Wiley StatsRef: Statistics Reference Online,* Wiley, New York. <https://doi.org/10.1002/9781118445112.stat06462.pub2>

McCormick, W.T., P.J. Schweitzer and T.W. White (1972): Problem decomposition and data reorganization by a clustering technique, *Operations Research,* **20**(5), 993--1009. <https://doi.org/10.1287/opre.20.5.993>

McInnes, L and Healy, J, UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction, ArXiv e-prints 1802.03426, 2018.

Sammon, J. W. (1969) A non-linear mapping for data structure analysis. *IEEE Trans. Comput.*, **C-18** 401--409.

Tenenbaum, J.B., de Silva, V. & Langford, J.C. (2000) A global network framework for nonlinear dimensionality reduction. *Science* **290**, 2319-2323.

Tsafrir, D., Tsafrir, I., Ein-Dor, L., Zuk, O., Notterman, D.A. and Domany, E. (2005): Sorting points into neighborhoods (SPIN): data analysis and visualization by ordering distance matrices, *Bioinformatics,* **21**(10) 2301--8. <https://doi.org/10.1093/bioinformatics/bti329>

van der Maaten, L.J.P.; Hinton, G.E. (Nov 2008). Visualizing Data Using t-SNE. *Journal of Machine Learning Research*. **9**: 2579--2605.
