---
title: "List of Seriation Criteria implemented in the R package seriation"
author: "Michael Hahsler"
format: 
  html:
    toc: true
editor: visual
---

This document contains the seriation criteria for judging the quality of permutations given data implemented in the R package [`seriation`](https://github.com/mhahsler/seriation).

This list was created for the following version of seriation:

```{r}
library("seriation")
packageVersion('seriation')
```

```{r, echo = FALSE}
.print_control <- function(control,
                           help = TRUE,
                           trim_values = 30L) {
  if (length(control) < 1L) {
    writeLines("no parameters")
  } else{
    contr <- lapply(
      control,
      FUN = function(x)
        strtrim(paste(deparse(x), collapse = ""), trim_values)
    )

    contr <- as.data.frame(t(as.data.frame(contr)))
    colnames(contr) <- "default"

    contr <- cbind(contr, help = "N/A")
    if (!is.null(attr(control, "help")))
      for (i in seq(nrow(contr))) {
        hlp <- attr(control, "help")[[rownames(contr)[i]]]
        if (!is.null(hlp))
        contr[["help"]][i] <- hlp
      }

    print(knitr::kable(contr))
  }

}

optimized_by <- function(criterion, add_link = TRUE) {
  l <- registry_seriate$get_entries()
  names <- sapply(l, "[[", "name")
  ops <- sapply(l, "[[", "optimizes")
  match <- ops == criterion
  match[is.na(match)] <- FALSE
  res <- unname(names[match])
  if (length(res) == 0) 
    return("N/A")
  
  .make_anchor <- function(x) {
    # remove leading numbers
    x <- gsub("^\\d+", "", x)
    # lower case
    tolower(x)
  }
  
  # add links 
  if (add_link)
    res <- sapply(res, FUN = function(m) paste0("[", m , "]", "(seriation_methods.html#", .make_anchor(m), ")"))
  
  res
}

print_criterion_method_md <- function(x, ...) {
    cat("\n\n###", x$name, "\n")
    cat("\n", x$description, "\n\n")
  
#    cat("* kind:", x$kind, "\n")
    cat("* merit:", x$merit, "\n")
    cat("* optimized by: ", paste(optimized_by(x$name), collapse = ", "), "\n")
    if (!is.na(x$registered_by)) {
      cat("* registered by: ", x$registered_by, "\n")
    }
    
    #    cat("* randomized:", x$randomized, "\n\n")
    

  writeLines("* additional parameters:")
  .print_control(x$control, trim_values = 100)

  
  invisible(x)
}

```

Register some additional criteria

```{r, results="hide", warning=FALSE, message=FALSE}
register_DendSer()
register_smacof()
```

The criteria are organized by methods that evaluate the permutation based on distance data or data matrices.

`merit` specified if the criteria function increases with better fit or if it is formulated as a loss criteria functions (`merit = FALSE`).

If a seriation method directly tries to optimize the criterion, than its name is specified under "optimized by". The names can be used as the `method` argument in `seriate()`.

## Criteria for dissimilarity data (dist)

```{r, results='asis', echo = FALSE}
l <- list_criterion_methods(kind = "dist", names_only = FALSE)
l <- l[order(sapply(l, "[[", "name"))]
for(i in seq_along(l))
  print_criterion_method_md(l[[i]])
```

## Criteria for data matrices, tables and data.frames

```{r, results='asis', echo = FALSE}
l <- list_criterion_methods(kind = "matrix", names_only = FALSE)
l <- l[order(sapply(l, "[[", "name"))]
for(i in seq_along(l))
  print_criterion_method_md(l[[i]])
```

# References

Barnard, S.T., A. Pothen, and H. D. Simon (1993): A Spectral Algorithm for Envelope Reduction of Sparse Matrices. *In Proceedings of the 1993 ACM/IEEE Conference on Supercomputing,* 493--502. Supercomputing '93. New York, NY, USA: ACM.

Caraux, G. and S. Pinloche (2005): Permutmatrix: A Graphical Environment to Arrange Gene Expression Profiles in Optimal Linear Order, *Bioinformatics,* **21**(7), 1280--1281.

Chen, C.-H. (2002): Generalized association plots: Information visualization via iteratively generated correlation matrices, *Statistica Sinica,* **12**(1), 7--29.

Deutsch, S.B. and J.J. Martin (1971): An ordering algorithm for analysis of data arrays. *Operational Research,* **19**(6), 1350--1362. \doi{10.1287/opre.19.6.1350}

Earle, D. and C.B. Hurley (2015): Advances in Dendrogram Seriation for Application to Visualization. *Journal of Computational and Graphical Statistics,* **24**(1), 1--25. \doi{10.1080/10618600.2013.874295}

Hahsler, M. (2017): An experimental comparison of seriation methods for one-mode two-way data. *European Journal of Operational Research,* **257**, 133--143. \doi{10.1016/j.ejor.2016.08.066}

Hubert, L. and J. Schultz (1976): Quadratic Assignment as a General Data Analysis Strategy. *British Journal of Mathematical and Statistical Psychology,* **29**(2). Blackwell Publishing Ltd. 190--241. \doi{10.1111/j.2044-8317.1976.tb00714.x}

Hubert, L., P. Arabie, and J. Meulman (2001): *Combinatorial Data Analysis: Optimization by Dynamic Programming.* Society for Industrial Mathematics. \doi{10.1137/1.9780898718553}

Niermann, S. (2005): Optimizing the Ordering of Tables With Evolutionary Computation, *The American Statistician,* **59**(1), 41--46. \doi{10.1198/000313005X22770}

McCormick, W.T., P.J. Schweitzer and T.W. White (1972): Problem decomposition and data reorganization by a clustering technique, *Operations Research,* **20**(5), 993-1009. \doi{10.1287/opre.20.5.993}

Robinson, W.S. (1951): A method for chronologically ordering archaeological deposits, *American Antiquity,* **16**, 293--301. \doi{10.2307/276978}

Tien, Y-J., Yun-Shien Lee, Han-Ming Wu and Chun-Houh Chen (2008): Methods for simultaneously identifying coherent local clusters with smooth global patterns in gene expression profiles, *BMC Bioinformatics,* **9**(155), 1--16. \doi{10.1186/1471-2105-9-155}
